How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-15 talk/dyla2007/dyla.tex
Shorten the abstract, trying to match more closely the content of the paper.

We argue in this paper that one should not write interpreters for dynamic
languages manually but rather use meta-programming techniques and raise the
overall level at which they are implemented. We believe this to be
ultimately a better investment of efforts than the development of more and more advanced
general-purpose object oriented virtual machines (VMs).

\medskip

Dynamic languages are traditionally implemented by writing a virtual machine
centered around an interpreter and/or a built-in compiler and providing
the object model and memory management. When a language becomes more popular,
the limitations of such an implementation lead to the emergence of alternative
implementations that try to solve some of the problems. Another reason for new
implementations is the desire to have the language integrate well with existing,
well-tuned object-oriented virtual machine like the Java Virtual Machine. In this paper, we
describe the mechanisms that lead to an abundance of implementations and
explore some of the limitations of standard VMs.  We propose a
complementary alternative to writing VMs by hand and dealing with low-level
details, validated by the PyPy project: flexibly generating virtual machines
from a single abstract language ``specification'', inserting features and
low-level details automatically -- including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research was partially supported by the EU funded
 project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details (like
garbage collection and threading issues). Limitations
of the C implementation lead to alternative implementations which draw
resources from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to the ability to re-use the libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters into conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand to have it run on various platforms -- high-level VMs as well as
C-level environments.

>>>>

Typical modern dynamic languages have a growing number of
implementations.  We explore the reasons for this situation, and the
limitations it imposes on open source or academic communities that lack
the resources to fine-tune and maintain them all.  It is sometimes
proposed that implementing dynamic languages on top of a standardized
general-purpose object-oriented virtual machine (like Java or .NET)
would help reduce this burden.  We propose a complementary alternative
to writing custom virtual machine (VMs) by hand, validated by the PyPy
project: flexibly generating VMs from a high-level ``specification'',
inserting features and low-level details automatically -- including good
just-in-time compilers tuned to the dynamic language at hand.  We
believe this to be ultimately a better investment of efforts than the
development of more and more advanced general-purpose object oriented
VMs.  In this paper we compare these two approaches in detail.
\footnote{This research was partially supported by the EU funded
 project: IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

For these reasons writing a virtual machine in C is problematic because it
forces the language implementer to deal with many low-level details (like
garbage collection and threading issues). Limitations
of the C implementation lead to alternative implementations which draw
resources from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM (Java Virtual Machine) or the
CLR (Common Language Runtime of the .NET framework). This is often wanted by
the community anyway, since it leads to the ability to re-use the libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters into conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand to have it run on various platforms -- high-level VMs as well as
C-level environments.
